\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-14 papers/jss/article.Rnw
Move the MapReduce / Histogram example before the OpenCPU one and greatly improve it.  Add a very simple python example of using the protoc and a few lines of python to store arrays of bin counts and breaks as a Histogram protocol buffer.
Then, use HistogramTools to read in this histogram into R, convert it
to a native R histogram object, and plot it.

Add a reference to blocker's theoretical work on preprocessing, and a
self-citation at the end of this example section to show a real
application of the design pattern described here.

TODO: All three steps - the python code, the R code, and the

%\section{Basic usage example - tutorial.Person}

@@

Over the past years, the MapReduce programming model \citep{dean2008mapreduce}
has emerged as a poweful design pattern for processing large data
sets in parallel on large compute clusters.  Protocol Buffers
provide a convenient mechanism to send structured data between tasks
in a MapReduce cluster.  In particular, the large data sets in fields
such as particle physics and information processing are frequently
stored in binned or histogram form in order to reduce the data storage
requirements for later data analysis \citep{scott2009multivariate}.

In such environments, analysts may be interested in very rare
phenomenon or be dealing with highly skewed data sets or inflexible
raw data storage systems from which unbiased sampling is not feasible.
There are two common patterns for generating histograms of large data
sets in a single pass with MapReduce.  In the first method, each
mapper task generates a histogram over a subset of the data that it
has been assigned, serializes this histogram and sends it to one or
more reducer tasks which merge the intermediate histograms from the
mappers.

In both methods, the mapper tasks must choose identical bucket
boundaries in advance if we are to construct the histogram in a single
pass, even though they are analyzing disjoint parts of the input set
that may cover different ranges.  The \pkg{HistogramTools} package
\citep{histogramtools} enhances \pkg{RProtoBuf} by providing a concise
schema for R histogram objects.  The histogram message type is
designed to be helpful if some of the Map or Reduce tasks are written
in R, or if those components are written in other languages and only
the resulting output histograms need to be manipulated in R.

>>>>

Many large data sets in fields such as particle physics and information
processing are stored in binned or histogram form in order to reduce
the data storage requirements \citep{scott2009multivariate}.  In the
last decade, the MapReduce programming model \citep{dean2008mapreduce}
has emerged as a popular design pattern that enables the processing of
very large data sets on large compute clusters.

Many types of data analysis over large data sets may involve very rare
phenomenon or be dealing with highly skewed data sets or inflexible
raw data storage systems from which unbiased sampling is not feasible.
In such situations, MapReduce and binning may be combined as a
pre-processing step for a wide range of statistical and scientific
analyses \citep{blocker2013}.

There are two common patterns for generating histograms of large data
sets in a single pass with MapReduce.  In the first method, each
mapper task generates a histogram over a subset of the data that it
has been assigned, serializes this histogram and sends it to one or
more reducer tasks which merge the intermediate histograms from the
mappers.

In both methods, the mapper tasks must choose identical bucket
boundaries in advance if we are to construct the histogram in a single
pass, even though they are analyzing disjoint parts of the input set
that may cover different ranges.  All distributed tasks involved in
the pre-processing as well as any downstream data analysis tasks must
share a schema of the histogram representation to coordinate
effectively.

The \pkg{HistogramTools} package \citep{histogramtools} enhances
\pkg{RProtoBuf} by providing a concise schema for R histogram objects:

\begin{example}
package HistogramTools;

message HistogramState {
  repeated double breaks = 1;
  repeated int32 counts = 2;
  optional string name = 3;
}
\end{example}

This HistogramState message type is designed to be helpful if some of
the Map or Reduce tasks are written in R, or if those components are
written in other languages and only the resulting output histograms
need to be manipulated in R.  For example, to create HistogramState
messages in Python for later consumption by R, we first compile the 
\texttt{histogram.proto} descriptor into a python module using the
\texttt{protoc} compiler:

\begin{verbatim}
  protoc histogram.proto --python_out=.
\end{verbatim}
This generates Python module called \texttt{histogram\_pb2.py}, containing both the 
descriptor information as well as methods to read and manipulate the R object 
message.

\begin{verbatim}
# Import modules
from histogram_pb2 import HistogramState;

# Create empty Histogram message
hist = HistogramState()

# Add breakpoints and binned data set.
hist.counts.extend([2, 6, 2, 4, 6])
hist.breaks.extend(range(6))
hist.name="Example Histogram Created in Python"

# Output the histogram
outfile = open("/tmp/hist.pb", "wb")
outfile.write(hist.SerializeToString())
outfile.close()
\end{verbatim}

We can then read in the histogram into R and plot it with :

\begin{verbatim}
library(RProtoBuf)
library(HistogramTools)

# Read the Histogram schema
readProtoFiles(package="HistogramTools")

# Read the serialized histogram file.
hist <- HistogramTools.HistogramState$read("/tmp/hist.pb")
hist
[1] "message of type 'HistogramTools.HistogramState' with 3 fields set"

# Convert to native R histogram object and plot
plot(as.histogram(hist))
\end{verbatim}

<<echo=FALSE,fig=TRUE>>=
require(RProtoBuf)
require(HistogramTools)
readProtoFiles(package="HistogramTools")
hist <- HistogramTools.HistogramState$read("/tmp/hist.pb")
plot(as.histogram(hist))
@

One of the authors has used this design pattern for several large
scale studies of distributed filesystems \citep{janus}.

@@
